---
title: "p8105_hw3_enp2116"
author: "Emily Potts"
date: "2022-10-06"
output: github_document
---

** This homework includes figures; the readability of your embedded plots (e.g. font sizes, axis labels, titles) will be assessed in Problems 1+ **


```{r, echo = FALSE, message = FALSE}
library(tidyverse)
library(ggridges)
library(patchwork)

library(p8105.datasets)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

# Problem 1

Loading the instacart dataset from p8105.datasets:
```{r, message = FALSE}
data("instacart")

instacart = 
  instacart %>% 
  as_tibble(instacart)
```


Exploration of the dataset:

The `instacart` dataset is tidy, with `r nrow(instacart)` rows and `r ncol(instacart)` columns, indicating we have data on  `r nrow(instacart)` different products ordered through instacart and we have information on `r ncol(instacart)` different dimensions in each of these product orders. The `instacart` dataset contains the variables: `r names(instacart)`. 

Key variables include user_id, order_id, and aisle. User_id has values such as `r head(unique(instacart$user_id))`, with `r instacart %>% select(user_id) %>% distinct %>% count` distinct users in total. Order_id such as `r head(unique(instacart$order_id))`, with `r instacart %>% select(user_id, order_id) %>% distinct %>% count` total orders in this dataset. Finally, aisle has values such as `r head(instacart$aisle)`. 

For a single order, there are multiple observations, such as all the products for order 1:
```{r, echo = FALSE, message = FALSE}
instacart %>%
 group_by(order_id) %>%
 filter(order_id == 1)
```


* How many aisles are there, and which aisles are the most items ordered from?
```{r, echo = FALSE, message = FALSE}
instacart %>% 
  distinct(aisle) %>% 
  nrow()

instacart %>% 
  group_by(aisle) %>% 
 summarize(
    n_obs = n()
  ) %>% 
  arrange(desc(n_obs))
```
There are 134 aisles. Fresh vegetables and fresh fruits are ordered most, with 150609 orders and 150473, respectively.

* Plot showing the number of items ordered in each aisle, limiting this to aisles with more than 10000 items ordered, and ordering aisles by number of orders.
```{r}
instacart %>%
  group_by(aisle) %>%
  summarize(
    n_obs = n()) %>% 
  filter(n_obs > 10000) %>% 
  mutate(aisle = forcats::fct_reorder(aisle, n_obs, desc)) %>% 
  ggplot(aes(x = aisle, y = n_obs)) + geom_point(aes(aisle, n_obs), alpha = .5) +
  labs(
    title = "Orders by Aisle",
    x = "Aisle Name",
    y = "Number of Orders"
    )  +  theme_bw() + theme(legend.position = "bottom") + theme(axis.text.x = element_text(angle = 90))

```

* Table showing the three most popular items in each of the aisles “baking ingredients”, “dog food care”, and “packaged vegetables fruits”, including the number of times each item is ordered.
```{r}
instacart %>% 
  filter(aisle == "baking ingredients" | aisle == "dog food care" | aisle == "packaged vegetables fruits") %>%
  group_by(aisle) %>% 
  count(product_name) %>% 
  mutate(rank = min_rank(desc(n))) %>% 
  filter(rank < 4) %>% 
  arrange(desc(n)) %>%
  knitr::kable()
```

* Table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week, formatted as a readable 2 x 7 table.
```{r, message = FALSE}
instacart %>%
  filter(product_name == "Pink Lady Apples" | product_name == "Coffee Ice Cream") %>%
  group_by(product_name, order_dow) %>%
  summarize(mean_hour = mean(order_hour_of_day)) %>%
  spread(key = order_dow, value = mean_hour) %>%
  knitr::kable(digits = 2)
```




# Problem 2

This problem uses five weeks of accelerometer data collected on a 63 year-old male with BMI 25, who was admitted to the Advanced Cardiac Care Center of Columbia University Medical Center and diagnosed with congestive heart failure (CHF). The data can be downloaded here. In this spreadsheet, variables activity.* are the activity counts for each minute of a 24-hour day starting at midnight.

* Loading, tidying, and otherwise wrangling the data; including all originally observed variables and values; have useful variable names; include a weekday vs weekend variable; and encode data with reasonable variable classes.

```{r, message = FALSE}
accel_data =
  read_csv("data/accel_data.csv") %>%
  janitor::clean_names() %>%
  pivot_longer(
    activity_1:activity_1440,
    names_to = "minute", 
    names_prefix = "activity_",
    values_to = "activity_count") %>%
  mutate(
    split_week = case_when(day == "Saturday" ~ 'weekend',
                          day == "Sunday" ~ 'weekend',
                          day == "Monday" ~ 'weekday',
                          day == "Tuesday" ~ 'weekday',
                          day == "Wednesday" ~ 'weekday',
                          day == "Thursday" ~ 'weekday',
                          day == "Friday" ~ 'weekday')) %>% 
  relocate(split_week) %>% 
  mutate(day = fct_relevel(day, c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")))
```

Describing the resulting dataset:

The `accel` dataset is tidy, with `r nrow(accel_data)` rows and `r ncol(accel_data)` columns, indicating we have accelerometer data for `r nrow(accel_data)` different minutes. The `accel` dataset contains variables: `r ls(accel_data)`. Day has values such as `r head(unique(accel_data$day))`, week such as `r head(unique(accel_data$week))`, and split_week such as `r head(unique(accel_data$split_week))`. For each single day, there are 1,440 activity counts, which are broken up by minute. Activity counts per minute take on values such as `r head(unique(accel_data$activity_count))`.


* Traditional analyses of accelerometer data focus on the total activity over the day. Using your tidied dataset, aggregate across minutes to create a total activity variable for each day, and create a table showing these totals. Are any trends apparent?

```{r, message = FALSE}
accel_data %>% 
  group_by(day, week) %>% 
  summarize(daily_total_activity = sum(activity_count)) %>% 
  pivot_wider(names_from = day, values_from = daily_total_activity) %>%
  knitr::kable(digits = 0)
```
Tuesday, Wednesday, and Thursdays all seem to have the consistent amounts of total activity (hovering around 300000-450000), whereas Fridays through Mondays are more variable (low of under 80000, high of almost 700000). These are expected patterns for weekdays versus weekends/extended weekends. A weird observation comes across in Saturday Weeks 4 & 5, where I predict that the accelerometer was not worn and the machine's default (NULL) is a 1 for each minute.

* Accelerometer data allows the inspection activity over the course of the day. Make a single-panel plot that shows the 24-hour activity time courses for each day and use color to indicate day of the week. Describe in words any patterns or conclusions you can make based on this graph.
```{r}
ggplot(accel_data, aes(x = minute, y = activity_count)) +
  geom_line(aes(color = day), alpha = .4) +
  labs(
    title = "Daily Activity by Minute",
    x = "Minute",
    y = "Activity Count"
    )  + scale_x_discrete(breaks = NULL)
```

Based on this graph, we can see that the distribution of activity throughout the day tends to be bimodal, with a lull before noon. On weekends, we see a larger nighttime peak and on weekdays we see a larger morning peak.


# Problem 3

Loading the NY NOAA dataset from p8105.datasets:
```{r, message = FALSE}
data("ny_noaa")

noaa_df = 
  ny_noaa %>% 
  as_tibble(ny_noaa)
noaa_df
```


A short description of the dataset, noting the size and structure of the data, describing some key variables, and indicating the extent to which missing data is an issue:

The `noaa_df` dataset has `r nrow(noaa_df)` rows and `r ncol(noaa_df)` columns, indicating we have weather data on  `r nrow(noaa_df)` different dates and we have information on `r ncol(noaa_df)` different dimensions, including date. The `noaa_df` dataset contains the variables: `r names(noaa_df)`. Prcp has values such as `r head(unique(noaa_df$prcp))`, snow such as `r head(unique(noaa_df$snow))`, and snwd such as `r head(unique(noaa_df$snwd))`. Finally, tmax and tmin are originally set up as character variables, which is not appropriate given they are number values and should be numeric. 

The `noaa_df` dataset does have a decent number of missing data (variables prcp, snow, snwd, tmax, tmin), most notably with almost 44% of temperature data missing. This missingness occurs primarily in the first half of the timeframe, but also in chunks in the second half, which may be an issue if missingness is not random (related to other variables in the dataset). The percentages for missing data are listed by variable:

```{r, echo = FALSE}
(colMeans(is.na(noaa_df)))*100
```

* Some data cleaning. Creating separate variables for year, month, and day & ensuring observations for temperature, precipitation, and snowfall are given in reasonable units:

To clean this data, I first transformed date into the variables of year, month, and day. I then converted the temperature variables to numeric and created new variables with my suggested units for the five variables of interest. Based on the documentation on the p8105 website for the NOAA dataset, I intended to have the variables of prcp, snow, and snwd in measures of cm and temperature data in terms of degrees C.
```{r}
noaa_tidy = 
  noaa_df %>%
  mutate(year = lubridate::year(date), 
                month = lubridate::month(date), 
                day = lubridate::day(date)) %>% 
  select(-date) %>% 
  mutate(tmax = as.numeric(tmax),
         tmin = as.numeric(tmin)) %>% 
  mutate(prcp_cm = round((prcp * 0.01), 1),
         snow_cm = round((snow * 0.1), 1),
         snwd_cm = round((snwd * 0.1), 1),
         tmax_C = round((tmax * 0.1), 1),
         tmin_C = round((tmin * 0.1), 1)) %>% 
    mutate(month = recode(month, "1" = "jan", "2" = "feb", "3" = "mar", "4" = "apr", "5" = "may", "6" = "jun", "7" = "jul", "8" = "aug", "9" = "sep", "10" = "oct", "11" = "nov", "12" = "dec"))
  
```

For snowfall, what are the most commonly observed values? Why?

For snowfall, the most commonly observed values are 0, with this value making up about 77.4% of the data. This is most likely the case because snow itself is a rare event, even in a cold climate that keeps snow on the ground. 
```{r}
noaa_tidy %>% 
  select(snow) %>% 
  filter(snow == 0) %>% 
  count() / 2595176
```

```{r}
noaa_tidy %>% 
  ggplot(aes(snow_cm)) + geom_histogram()
```


* Make a two-panel plot showing the average max temperature in January and in July in each station across years. Is there any observable / interpretable structure? Any outliers?

```{r}
  noaa_tidy %>% 
  group_by(id, month, year) %>% 
  summarize(
    mean_tmax = mean(tmax_C)
  ) %>%
  filter(month == "jan" | month == "jul") %>%
  ggplot(aes(x = year, y = mean_tmax, color = id)) + 
  geom_line() +   theme(legend.position="none") + labs(x = "Years", y = "Average Max Temperature (degrees C)") + facet_grid(~ month)

```
Based on our two-panel plot, we can see a clear seasonal trend for average max temperature when comparing january and july; january has hovered around 0 degrees C while July around 27.5 degrees C. Additionally, at a given time point, different stations seem to mirror one anothers' trends (within about 10 degrees C). Finally, over the 30 year timespan of interest, average max temperature has not changed much when comparing within the same months.

* Make a two-panel plot showing (i) tmax vs tmin for the full dataset (note that a scatterplot may not be the best option); and (ii) make a plot showing the distribution of snowfall values greater than 0 and less than 100 separately by year.

- You should make a single plot for part (i) and a single plot for part (ii). Then, you should combine these two plots side by side into a single-panel exhibit; you can find a refresher on how to do that hereLinks to an external site.. For the first plot, you are correct that "vs" implies tmax should be on the vertical axis and that tmin should be on the horizontal axis.

- As for the final plot, you should filter values based on the original definition of snowfall (i.e., using the original units). To that end, you might want to consider retaining both the original variable snow as well as your proposal for a variable with a more sensical unit. As long as you're clear about how you label values and units in your visualizations, you should be good!

- filter the data to only include rows such that 0 < snow < 100.

- Then for each year of the data, create distribution plots of snow. So, for example, if the dataset spanned the years 1997 to 1999, you would have three separate distribution plots of (the filtered observations of) snow by year.

- patchwork, geom_hex

- We want to limit the values of snowfall to the criteria specified and then use ggplot to plot the distribution (density plot). 
